Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-04-13 2020/RSD_GPU/RSD2020.tex
Обновленные таблицы. Поправлено оформление.

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA); (results of the first two algorithms are given in the article \cite{Sergeyev2006}). A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Sergeyev2006}. The Global minimum $y*$ was considered found if the algorithm generated a test point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\|\le \delta $. Also, the size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved, a и b are the boundaries of the search domain $D$, the parameter  $\Delta $=10-6 for N=4 and $\Delta =10-7$ for $N=5$. When using the AGP method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class - $r=5.6$; the parameter for constructing the Peano curve was fixed, $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & DIRECT & DIRECT\(l\) & GSA \\ \hline
    4 & Simple & >47282 (4) & 18983 & 11953\\ \cline{2-5}
      & Hard & >95708 (7) & 68754 & 25263\\ \hline
    5 & Simple &>16057 (1) & 16758 & 15920\\ \cline{2-5}
      & Hard & >217215 (16) & >269064 (4) & >148342 (4)\\ \hline
    \end{tabular}
    \caption{Average number of iterations}
    \label{table:average_iters}
\end{table}

   Table \ref{table:average_iters} shows the average number of iterations, $k_{av}$, that the method performed when solving a series of problems from these classes. The symbol ">" reflects a situation where not all problems of the class were solved by any method whatsoever. This means that the algorithm was halted because the maximum allowed number of $K_{max}$ iterations was reached. In this case, the value $K_{max} = 10^6$ was used to calculate the average value of the number of iterations, $k_{av}$, which corresponds to the lower estimate of this average value. The number of unsolved problems is indicated in parentheses.

@@

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple  & 3,49 & 8,22  &  \\ \cline{2-5}
      & Hard    & 2,35 & 7,65  &  \\ \hline
    5 & Simple  & 1,87 & 6,59  &  \\ \cline{2-5}
      & Hard    & 1,48 & 10,80  &   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the CPU}
    \label{table:average_iters_AS_CPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple  & 1,71 & 2,01  &  \\ \cline{2-5}
      & Hard    & 1,17 & 1,90  &  \\ \hline
    5 & Simple  & 0,88 & 1,48  &  \\ \cline{2-5}
      & Hard    & 0,70 & 2,39  &   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the CPU}
    \label{table:average_Time_AS_CPU}
\end{table}

   Let us now evaluate the acceleration achieved using parallel GSA using an adaptive dimensionality reduction scheme based on the number $P$ of cores used. Table \ref{table:average_iters_AS_CPU} shows the acceleration by iterations, and Table \ref{table:average_Time_AS_CPU} shows the time acceleration obtained when solving a series of problems on the CPU, compared to GSA.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple &   6,62 & 21,46  &  \\ \cline{2-5}
      & Hard   &   3,23 & 9,97  &  \\ \hline
    5 & Simple &   3,37 & 11,20  &  \\ \cline{2-5}
      & Hard   &   2,52 & 8,54  &   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the CPU}
    \label{table:average_iters_MMAS_CPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=2 & P=4 & P=8 \\ \hline
    4 & Simple &   3,22 & 5,23  &  \\ \cline{2-5}
      & Hard   &   1,48 & 2,42  &  \\ \hline
    5 & Simple &   1,30 & 1,97  &  \\ \cline{2-5}
      & Hard   &   0,68 & 0,88  &   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the CPU}
    \label{table:average_Time_MMAS_CPU}
\end{table}

   Table \ref{table:average_iters_MMAS_CPU} shows the acceleration in iterations, and Table \ref{table:average_Time_MMAS_CPU} shows the acceleration in time of the algorithm that combines multiple sweeps and an adaptive scheme for solving a series of problems on the CPU, compared to the sequential launch of the GSA. Two scans and, accordingly, two processes were used; each process used $P$ threads, and calculations were performed on a single cluster node.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   5,06 & 19,91 & 15,62  \\ \cline{2-5}
      & Hard   &   3,93 & 14,95 &   \\ \hline
    5 & Simple &   4,16 & 16,12 & 67,22  \\ \cline{2-5}
      & Hard   &   3,36 & 10,02 & 42,04   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the GPU}
    \label{table:average_iters_AS_GPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   2,20 & 8,28 & 5,21  \\ \cline{2-5}
      & Hard   &   1,92 & 6,91 &   \\ \hline
    5 & Simple &   1,58 & 4,18 & 8,23  \\ \cline{2-5}
      & Hard   &   1,22 & 2,39 & 3,74   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the GPU}
    \label{table:average_Time_AS_GPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   11,32 & 38,96 & 128,39  \\ \cline{2-5}
      & Hard   &   6,83  & 31,52 &   \\ \hline
    5 & Simple &   4,99  & 39,37 & 80,56  \\ \cline{2-5}
      & Hard   &   4,61  & 17,28 & 113,61   \\ \hline
    \end{tabular}
    \caption{Acceleration of iterations on the GPU}
    \label{table:average_iters_MMAS_GPU}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & P=64 & P=256 & P=1024 \\ \hline
    4 & Simple &   2,44 & 8,01 & 20,45  \\ \cline{2-5}
      & Hard   &   1,63 & 7,04 &   \\ \hline
    5 & Simple &   0,77 & 3,97 & 2,37  \\ \cline{2-5}
      & Hard   &   0,61 & 0,96 & 2,11   \\ \hline
    \end{tabular}
    \caption{Time acceleration on the GPU}
    \label{table:average_Time_MMAS_GPU}
\end{table}

Tables \ref{table:average_iters_AS_GPU} and  \ref{table:average_Time_AS_GPU} show the acceleration obtained when solving a series of problems on the GPU using an adaptive scheme compared to a similar launch on the CPU using 4 threads. Tables \ref{table:average_iters_MMAS_GPU} and \ref{table:average_Time_MMAS_GPU} show the acceleration of the algorithm combining multiple sweeps and an adaptive circuit when solving a series of problems on the GPU, compared to an adaptive scheme on the CPU using 4 threads. Two sweeps and, accordingly, two processes were used; each process used P threads, and calculations were performed on a single cluster node.

>>>>

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA); (results of the first two algorithms are given in the article \cite{Sergeyev2006}).  A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Sergeyev2006}. The Global minimum $y*$ was considered found if the algorithm generated a test point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\|\le \delta $.  Also, the size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved , a and b are the boundaries  of the search domain $D$, the parameter  $\Delta ={{10}^{-6}}$ for N=4 and $\Delta ={{10}^{-7}}$ for $N=5$. When using the AGP method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class - $r=5.6$; the parameter for constructing the Peano curve was fixed, $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & DIRECT & DIRECT\(l\) & GSA \\ \hline
    4 & Simple & \(>\)47282 (4) & 18983 & 11953\\ \cline{2-5}
      & Hard & \(>\)95708 (7) & 68754 & 25263\\ \hline
    5 & Simple &\(>\)16057 (1) & 16758 & 15920\\ \cline{2-5}
      & Hard & \(>\)217215 (16) & \(>\)269064 (4) & \(>\)148342 (4)\\ \hline
    \end{tabular}
    \caption{Average number of iterations}
    \label{table:average_iters}
\end{table}

   Table \ref{table:average_iters} shows the average number of iterations, $k_{av}$, that the method performed when solving a series of problems from these classes. The symbol \(">"\) reflects a situation where not all problems of the class were solved by any method whatsoever. This means that the algorithm was halted because the maximum allowed number of $K_{max}$ iterations was reached. In this case, the value $K_{max} = 10^6$ was used to calculate the average value of the number of iterations, $k_{av}$, which corresponds to the lower estimate of this average value. The number of unsolved problems is indicated in parentheses.

@@

\begin{table}
                \caption{Results on CPU}
                \label{table:average_AS_CPU}
                \center
                \begin{tabular}{cccccccccccc}
                               \hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Acceleration of iterations} & & \multicolumn{5}{c}{ Time acceleration}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               $p$ & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=2$&   3,49 & 2,35 &   & 0,86 & 2,03 &   & 1,71 & 1,17 &   & 0,95 & 0,4 \\
                               $p=4$&   8,22 & 7,65 &   & 6,74 & 8,56 &   & 2,01 & 1,9  &   & 1,85 & 1,53 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

   Let us now evaluate the acceleration achieved using parallel GSA using an adaptive dimensionality reduction scheme based on the number $P$ of cores used. Table \ref{table:average_AS_CPU} shows the acceleration by iterations and the time acceleration obtained when solving a series of problems on the CPU, compared to GSA.

\begin{table}
                \caption{Results on CPU}
                \label{table:average_MMAS_CPU}
                \center
                \begin{tabular}{cccccccccccc}
                               \hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Acceleration of iterations} & & \multicolumn{5}{c}{ Time acceleration}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               $p$ & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=2$&    6,62  & 3,23  &   & 2,13 & 6,07  &   & 3,22 & 1,48 &   & 0,66 & 1,93 \\
                               $p=4$&    21,46 &  9,97 &   & 6,9  & 16,27 &   & 5,23 & 2,42 &   & 0,85 & 1,9 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

   Table \ref{table:average_MMAS_CPU} shows the acceleration in iterations and  the acceleration in time of the algorithm that combines multiple sweeps and an adaptive scheme for solving a series of problems on the CPU, compared to the sequential launch of the GSA. Two scans and, accordingly, two processes were used; each process used $P$ threads, and calculations were performed on a single cluster node.

\begin{table}
                \caption{Results on GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{cccccccccccc}
                               \hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Acceleration of iterations} & & \multicolumn{5}{c}{ Time acceleration}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               $p$ & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,06  & 3,93  &   & 1,19 	 & 5,41  &   & 2,2	    & 1,92	 &   & 0,47	& 2,21 \\
                               $p=256$ &    19,91 & 14,95 &   & 11,9 	 & 23,36 &   & 8,28	& 6,91	     &   & 3,4	    & 6,1 \\
							   $p=1024$&    15,62 & 52,91 &   & 22,12	 & 55,22 &   & 5,21	& 19,95      &   & 2,18	& 4,82 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Results on GPU}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{cccccccccccc}
                               \hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Acceleration of iterations} & & \multicolumn{5}{c}{ Time acceleration}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               $p$ & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & 6,83	 &   & 8,19	     & 10,66  &   & 2,44	 & 1,63	 &   & 1,33	 & 1,58 \\
                               $p=256$ &    38,96	 & 31,52 &   & 25,43	 & 48,23  &   & 8,01	 & 7,04	 &   & 2,31	 & 3,42 \\
							   $p=1024$&    128,39	 & 83,46 &   & 98,43	 & 146,23 &   & 20,45	 & 11,06 &   & 2,75	 & 2,48 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

Tables \ref{table:average_AS_GPU} show the acceleration obtained when solving a series of problems on the GPU using an adaptive scheme compared to a similar launch on the CPU using 4 threads. Tables \ref{table:average_MMAS_GPU} show the acceleration of the algorithm combining multiple sweeps and an adaptive circuit when solving a series of problems on the GPU, compared to an adaptive scheme on the CPU using 4 threads. Two sweeps and, accordingly, two processes were used; each process used $P$ threads, and calculations were performed on a single cluster node.
